### 2025-11-05 sections/3-datasets.tex
removed another half a column

Table~\ref{tab:datasets:summary} summarizes the cardinality and dimensionality of each dataset we use.
These include the Fashion-MNIST, Glove-25, and Sift datasets from the ANN-Benchmarks suite~\cite{aumuller2020ann}, as well as a Random dataset we generated (using the same cardinality and dimensionality as Sift) as an anti-example to demonstrate the effect of manifold structure on CAKES performance.
Thus, Random shows that CAKES is designed to exploit manifold structure when it exists, and should not be used on datasets that do not obey the manifold hypothesis.
The Glove-25 dataset uses the Cosine distance function, while the other three datasets use the Euclidean distance metric.
All benchmarks using these datasets were conducted on an r6i.16xlarge instance on Amazon Web Services (AWS), with 64 vCPUs and 512 GiB of RAM.
The Rust compiler was Rust 1.83.0 and the Python interpreter version was 3.9.18.

We also include benchmarks on the SilVA-18S rRNA and Radio-ML datasets.
For the SilVA-18S dataset, which consists of 2,224,640 sequences of up-to 3,712 nucleotides, we use the Levenshtein distance metric~\cite{levenshtein1966binary}.
The Radio-ML dataset consists of 1,024-dimensional complex-valued time-series samples, spanning 24 modulation modes and 26 SNR levels from $-20$dB to $30$dB.
We use the 10dB SNR subset with 97,304 samples and use Dynamic Time Warping~\cite{muller2007dynamic} as the distance metric.
For both datasets, we hold out 1,000 random samples as queries and build the tree on the remaining samples.
These were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM running Manjaro Linux 5.15.164-1-MANJARO with Rust compiler version 1.83.0.

@@

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axis. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD indicating the presence of manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process that imparts some manifold structure at three distinct radial scales, indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

@@

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} show the throughput (QPS) and recall for each algorithm on each of the vector datasets.
Since HNSW, ANNOY, and FAISS do not support the Levenshtein or DTW distance functions, we only benchmark CAKES on the SilVA-18S and RadioML datasets and report results in Figures~\ref{fig:results:silva-scaling} and~\ref{fig:results:radioml-scaling}.
All benchmarks were run with $k=10$, the throughput is averaged over all queries, and recall is computed using a na\"ive linear scan.

HNSW, ANNOY, and FAISS have various hyper-parameters that can be tuned to trade off throughput, recall, and indexing time.
We allow for a hyper-parameter search (in the ranges recommended by the documentation accompanying the respective libraries) to find the highest throughput configuration that achieves perfect recall (or as close to it as possible) for each dataset and cardinality.
To make the comparison fair, we also allow CAKES to select the optimal algorithm among the three CAKES algorithms for each dataset and cardinality by using a small number of the centers of clusters at low depth as representative queries.

>>>>

Table~\ref{tab:datasets:summary} summarizes the cardinality and dimensionality of each dataset we use.
These include the Fashion-MNIST, Glove-25, and Sift datasets from the ANN-Benchmarks suite~\cite{aumuller2020ann}, as well as a Random dataset we generated (using the same cardinality and dimensionality as Sift) as an anti-example to demonstrate the effect of manifold structure on CAKES performance.
The Glove-25 dataset uses the Cosine distance function, while the other three datasets use the Euclidean distance metric.
All benchmarks using these datasets were conducted on an r6i.16xlarge instance on Amazon Web Services (AWS), with 64 vCPUs and 512 GiB of RAM.
The Rust compiler was Rust 1.83.0 and the Python interpreter version was 3.9.18.

We also include benchmarks on the SilVA-18S and Radio-ML datasets.
For the SilVA-18S dataset, which consists of 2,224,640 sequences of up-to 3,712 nucleotides, we use the Levenshtein distance metric~\cite{levenshtein1966binary}.
The Radio-ML dataset consists of 1,024-dimensional complex-valued time-series samples, spanning 24 modulation modes and 26 SNR levels from $-20$dB to $30$dB.
We use the 10dB SNR subset with 97,304 samples and Dynamic Time Warping~\cite{muller2007dynamic} as the distance metric.
For both datasets, we hold out 1,000 random samples as queries and build the tree on the remaining samples.
These were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM running Manjaro Linux 5.15.164-1-MANJARO.

@@

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axis. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD indicating the presence of manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process. It seems to have little manifold structure, if any, instead exhibiting behavior akin to the Random dataset across three separate radial scales, indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

@@

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} show the throughput (QPS) and recall for each algorithm on each of the vector datasets.
Since HNSW, ANNOY, and FAISS do not support the Levenshtein or DTW distance functions, we only benchmark CAKES on the SilVA-18S and RadioML datasets and report results in Figures~\ref{fig:results:silva-scaling} and~\ref{fig:results:radioml-scaling}.
All benchmarks were run with $k=10$, throughput is averaged over all queries, and recall is computed using a na\"ive linear scan.

HNSW, ANNOY, and FAISS have various hyper-parameters that can be tuned to trade off throughput, recall, and indexing time.
We allow for a hyper-parameter search (in the ranges recommended by the documentation accompanying the respective libraries) to find the configuration with the highest throughput that still achieves perfect recall (or as close to it as possible) for each dataset and cardinality.
To make the comparison fair, we also allow CAKES to select the fastest of its algorithms for each dataset and cardinality by testing each with a small number of the centers of clusters at low depth as representative queries.

### 2025-11-05 sections/4-results.tex
removed another half a column

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} summarize the throughput and recall of CAKES against state-of-the-art $k$-NN search algorithms on the Fashion-MNIST, Glove-25, Sift, and Random datasets respectively.
Throughput is measured in queries per second (QPS), and a recall of $1.000*$ indicates imperfect recall that rounds to $1.000$.

We see that CAKES has perfect recall on all datasets using the Euclidean distance metric, and near-perfect recall on the Glove-25 dataset using Cosine distance.
This is because Cosine distance violates the triangle inequality and is not a metric.
The throughput of CAKES drops slightly for the first two multipliers, but then remains relatively constant as cardinality increases, demonstrating the scalability of CAKES on manifold-structured data.
In contrast, the recall of HNSW and ANNOY degrades significantly as cardinality increases, even though their throughput remains high.
FAISS-IVF keeps high recall but its throughput degrades linearly with cardinality.
This indicates that these approximate algorithms are unable to maintain both high recall and high throughput as dataset sizes increase, even on manifold-structured data.

For the Random dataset, CAKES again attains perfect recall at all cardinalities, while HNSW and ANNOY exhibit near-zero recall, and FAISS-IVF attains perfect recall but with very low throughput.

On the Silva dataset (Fig.~\ref{fig:results:silva-scaling}), all algorithms initially exhibit throughput that decreases approximately linearly with cardinality, but then level off at higher cardinalities.
CAKES's Depth-First Sieve is consistently the fastest algorithm on this dataset.

On the Radio-ML dataset (Fig.~\ref{fig:results:radioml-scaling}), all CAKES variants are slower than na\"{i}ve linear search, and their throughput decreases linearly with cardinality.
This is expected, as the Radio-ML dataset is a randomly generated dataset and exhibits little manifold structure under the DTW distance metric.

\subsection{Indexing and Tuning Time}
\label{sec:results:indexing-and-tuning-time}

@@

  \vskip -0.1in
  \subfloat[Sift for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/sift-indexing.png}
    \label{fig:results:sift-indexing}
  }
  \subfloat[Random dataset for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/random-indexing.png}
    \label{fig:results:random-indexing}
  }
  \caption{Indexing and tuning time.
  We report the time taken by various algorithms to build and tune its index on various datasets.
  For CAKES, this includes the time required to build  the tree and perform auto-tuning.
  All axes are on logarithmic scales.
  For all datasets, we see that the indexing time for CAKES increases roughly linearly with the cardinality of the dataset.
  HNSW and ANNOY are the slowest at indexing.
  FAISS-Flat does not actually build an index;
  it provides a na\"{i}ve linear search in the FAISS family of algorithms.
  It is included only to illustrate the base cost of simply having a dataset.}
  \label{fig:results:indexing}
\end{figure}

>>>>

Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, and \ref{tab:results:qps-and-recall-random} summarize the throughput and recall of CAKES against state-of-the-art $k$-NN search algorithms on the Fashion-MNIST, Glove-25, Sift, and Random datasets respectively.
A recall of $1.000*$ indicates imperfect recall that rounds to $1.000$.

We see that CAKES has perfect recall on all datasets using the Euclidean distance metric, and near-perfect recall on the Glove-25 dataset using Cosine distance.
This is because Cosine distance violates the triangle inequality and is not a metric.
The throughput of CAKES drops slightly for the first two multipliers, but then remains relatively constant as cardinality increases, demonstrating the scalability of CAKES on manifold-structured data.
In contrast, the recall of HNSW and ANNOY degrades significantly as cardinality increases, even though their throughput remains high.
FAISS-IVF keeps high recall but its throughput degrades linearly with cardinality.
For the Random dataset, CAKES again attains perfect recall at all cardinalities, while HNSW and ANNOY exhibit near-zero recall, and FAISS-IVF attains perfect recall but with very low throughput.

On Silva (Fig.~\ref{fig:results:silva-scaling}), all CAKES algorithms initially exhibit throughput that decreases approximately linearly with cardinality, but then level off at higher cardinalities.
DFS is consistently the fastest algorithm on this dataset.

On Radio-ML (Fig.~\ref{fig:results:radioml-scaling}), all CAKES variants are slower than na\"{i}ve linear search, and their throughput decreases linearly with cardinality.

\subsection{Indexing Time}
\label{sec:results:indexing-and-tuning-time}

@@

  \vskip -0.1in
  \subfloat[Sift for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/sift-indexing.png}
    \label{fig:results:sift-indexing}
  }
  \subfloat[Random dataset for $k=10$.]{
    \includegraphics[width=0.24\textwidth]{images/indexing/random-indexing.png}
    \label{fig:results:random-indexing}
  }
  \caption{Indexing and tuning time.
  We report the time taken by various algorithms to build and tune its index on various datasets.
  For CAKES, this includes the time required to build the tree and find the fastest internal algorithm.
  All axes are on logarithmic scales.
  For all datasets, we see that the indexing time for CAKES increases roughly linearly with the cardinality of the dataset.
  HNSW and ANNOY are the slowest at indexing.
  FAISS-Flat does not actually build an index;
  it provides a na\"{i}ve linear search in the FAISS family of algorithms.
  It is included only to illustrate the base cost of simply having a dataset.}
  \label{fig:results:indexing}
\end{figure}

### 2025-11-05 sections/5-discussion.tex
removed another half a column

We have presented CAKES, a suite of three algorithms for fast $k$-NN search across diverse distance functions.
When the distance is a metric (Sec.~\ref{sec:methods}), CAKES is exact.
Under Cosine distance, a non-metric, CAKES achieves near-perfect recall.
Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove} and \ref{tab:results:qps-and-recall-sift}, and Figure~\ref{fig:results:silva-scaling} show that CAKES is most effective when data satisfy the manifold hypothesis, i.e., they occupy a low-dimensional manifold within a high-dimensional space.
Conversely, performance degrades on randomly generated data, where such structure is absent, as shown in Table~\ref{tab:results:qps-and-recall-random} and Figure~\ref{fig:results:radioml-scaling}.

In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.
As expected, CAKES scales sub-linearly with cardinality on real-world datasets with low LFD and linearly on synthetic datasets with high LFD;
Sift versus Random illustrates this contrast, as does Radio-ML.

>>>>

We have presented CAKES, a suite of three algorithms for fast $k$-NN search across diverse distance functions.
When the distance is a metric (Sec.~\ref{sec:methods}), CAKES is exact.
Even under a non-metric, namely Cosine distance, CAKES achieves near-perfect recall.
Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove} and \ref{tab:results:qps-and-recall-sift}, and Figure~\ref{fig:results:silva-scaling} show that CAKES is most effective when data satisfy the manifold hypothesis, i.e., they occupy a low-dimensional manifold within a high-dimensional space.
Conversely, performance degrades on randomly generated data, where such structure is absent, as shown in Table~\ref{tab:results:qps-and-recall-random} and Figure~\ref{fig:results:radioml-scaling}.
HNSW and ANNOY do maintain high throughput but their recall degrades significantly as cardinality grows, and is near zero on the Random dataset.
FAISS-IVF maintains high recall but its throughput degrades significantly as cardinality grows.

In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset exhibits a manifold structure, as quantified by LFD.
As expected, CAKES scales sub-linearly with cardinality on real-world datasets with low LFD and linearly on synthetic datasets with high LFD;
Sift versus Random illustrates this contrast, as does Radio-ML.
